---
title: "Geog6300: Three tests of normality"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
library(tidyverse)
```

This script covers three ways to assess the normality of a dataset. First, let's create two sample distributions. The first two are "beta" distribution, where the first is negatively skewed and the second is positively skewed. The third is a normal distribution.

```{r}
test1<-data.frame(var1=rbeta(5000,7,2))
test2<-data.frame(var2=rbeta(5000,1,5))
test3<-data.frame(var3=rnorm(5000,0,1))
```

Histograms show this difference pretty clearly.

```{r}
ggplot(test1,aes(x=var1)) + 
  geom_histogram()

ggplot(test2,aes(x=var2)) + 
  geom_histogram()

ggplot(test3,aes(x=var3)) + 
  geom_histogram()

```

We can also use a qqplot to assess normality. The "qq line" shows what this plot would look like if the distribution is normal.

```{r}
ggplot(test1,aes(sample=var1)) +
  stat_qq()+
  stat_qq_line()

ggplot(test2,aes(sample=var2)) +
  stat_qq()+
  stat_qq_line()

ggplot(test3,aes(sample=var3)) +
  stat_qq()+
  stat_qq_line()

```

The Shapiro-Wilk test is a more rigorous way to assess the normality of your data. The null hypothesis is that the data ARE normal, so a low p value indicates that the dataset is skewed/not normal.

```{r}
shapiro.test(test1$var1) 
shapiro.test(test2$var2) 
shapiro.test(test3$var3) 
```

### You try it!
Go back to our census data (ACSCtyData_2022_ACS). What's the most normal variable you can find?

```{r}
acsdata<-read_csv("data/ACSCtyData_2022ACS.csv")


```

