---
title: "Geog4/6300-Working with raster data"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

## Working with raster data in R

The raster package in R provides a suite of tools for loading and working with raster data. We can load a version of the National Land Cover Database for Houston as a geotiff using the `rast` function from the terra package as shown below.

As a reminder, a raster consists of a grid of equally sized cells. Each cell stores a value. In the NLCD dataset, these values represent land cover classes, while in a DEM they represent elevation.

```{r}
#install.packages("raster")
library(terra)
library(sf)
library(tidyverse)
library(tidycensus)
library(FedData)
library(tmap)

seattle<-st_read("data/us_2025_uac20.gpkg") %>%
  filter(str_detect(NAME20,"Seattle")) %>%
  select(NAME20) %>%
  st_transform(4326)

nlcd_legend <- pal_nlcd() %>%
  mutate(ID=as.character(ID))

nlcd25<-rast("data/nlcd_conus_2025_seattle.tif")
plot(nlcd25,col=nlcd_legend$Color)
```

The NLCD is a categorical raster--each cell belongs to a discrete land cover class (forest, developed land, wetlands, etc.) rather than storing a continuous numeric measurement, so you could look at the number of pixels/cells in each category using the `freq` function. In the code below, this is wrapped in a data frame so that it comes out as usable table.

```{r}
freq_table<-data.frame(freq(nlcd25))

#Remove the black/0 value cells and add in a description of each land use type
freq_table_join<-freq_table %>%
  filter(value!=0) %>%
  mutate(value=as.character(value)) %>% #Treats the numeric land use code as a character
  left_join(nlcd_legend %>% rename(value=ID))
```

## You try it!
Create a new column called percent_land showing the percentage of all non-zero NLCD cells represented by each land-cover class. Which class occupies the largest percentage of land area? You'd need to get the total number of pixels and then divide the counts for each area by that number.

```{r}

```

# Value extraction

What if you have point data and want to extract values for the raster? Let's load points for Starbucks in Houston. The extract function (from raster) pulls the value for each point. What's the most common land use type for each one?

```{r}
starbucks_seattle<-read_csv("data/brandlocations.csv") %>%
  filter(city=="Seattle" & region=="WA" & brands=="Starbucks") %>%
  st_as_sf(coords=c("X","Y"),crs=4326) %>%
  st_transform(st_crs(nlcd25))

starbucks_nlcd<-terra::extract(nlcd25,starbucks_seattle)
```

The resulting object is just a simple vector of elevation values. We have to bind it back to the original data frame. The easiest way to do so is using base R.

```{r}
starbucks_seattle$nlcd<-starbucks_nlcd

View(starbucks_seattle)
```

We can also plot these values using tmap.

```{r}
tm_shape(starbucks_seattle)+
  tm_dots("nlcd")
```

## You try it!
Create a frequency table showing the number of Starbucks locations in each NLCD class. Which land-cover type contains the most Starbucks locations?

```{r}

```

We can also use a different package, `exactextractr`, to identify the most common land use type in a subarea. Here, we create a simple hex cell grid using `st_make_grid`.

You can see more of what's possible with this package at this link: https://cran.r-project.org/web/packages/exactextractr/vignettes/vig2_categorical.html

```{r}
library(exactextractr)
nlcd_grid<-st_make_grid(nlcd21,cellsize = 10000) %>% st_as_sf()

extracted_values <- exact_extract(nlcd25, nlcd_grid,'mode') 

nlcd_grid$ID<-extracted_values

nlcd_grid<-nlcd_grid %>%
  left_join(legend)

ggplot(nlcd_grid)+
  geom_sf(aes(fill=Class))
```

## You try it!
Repeat the analysis using either a smaller grid size (5 km) or a larger grid size (20 km). How does changing the aggregation scale affect the spatial pattern?

```{r}

```

#Working with elevation data
The NLCD is categorical data. What about continuous/ratio data like elevation? We can load a digital elevation model (DEM) dataset for the Houston area with the raster package.

```{r}
#Unzip the dem if necessary
#unzip("data/dem_houston.zip",exdir="data")

houston_dem<-raster("data/dem_houston.tif")
plot(houston_dem)
```

What if we wanted to get mean values of this elevation data by area? This is called zonal statistics. We often use this with census geographies to be able to link raster data to demographics characteristics. In this case, we'll just make a simple hexagonal grid to aggregate these data.

The `exact_extract` function can also be used to calculate the mean elevation per cell.

```{r}
library(terra)
houston_bbox=st_bbox(houston_dem) %>% st_as_sfc()

houston_grid<-st_make_grid(houston_bbox,cellsize=0.05,square=FALSE) %>%
  st_as_sf() %>%
  mutate(id=row_number())
plot(houston_grid)

dem_extract<-exact_extract(houston_dem,houston_grid,"mean")

houston_grid1<-houston_grid %>%
  bind_cols(data.frame(dem_houston=dem_extract)) %>%
  filter(is.na(dem_houston)==FALSE)

``` 
 
It looks like grid cell 407 has the highest elevation. Let's map that out.

```{r}
tm_shape(houston_grid1)+
  tm_polygons("dem_houston",style="jenks")


ggplot(houston_grid1)+
  geom_sf(aes(fill=dem_houston))
```

## Mapping countours and slope

We can use the `contour` function to add contours to a plot of the elevations. The `add=TRUE` parameter keeps the underlying map.

```{r}
plot(houston_dem)
contour(houston_dem,add=TRUE,col="brown")
```

The terrain function allows you to extract properties like slope and aspect. Here, we map both slope and then select only those areas with a slope greater than 10 degrees. The third line of code is a raster reclassification, making a raster where 1 represents cells that meet the criteria (slope > 10%) and 0 is those cells that don't.

```{r}
x <- terrain(houston_dem, opt = "slope", unit = "degrees")
plot(x$slope)

x1<-x >= 10

plot(x1)
```

Note that raster is an older package, and both `terra` (created as the successor to `raster`) and `stars` are newer. 

Here's another walkthrough of raster data in R with some more advanced functions:
https://ourcodingclub.github.io/tutorials/spatial/

For 3D Visualization, the rayshader package can do some pretty amazing things. Here's one walkthrough: https://wcmbishop.github.io/rayshader-demo.
