---
title: "Filter, group_by, and summarise"
format: html
---

## Filtering data

What if you only want data for some states, like the Southeast?
Or counties with >30% poverty?
You can use "filter" (part of the tidyverse).
Remember that the tidyverse uses the pipe, which looks like this: %>%

```{r}
census_data_se<-census_data %>%
  filter(region=="SE")
```

What states are in our new data?
```{r}
table(census_data_se$st_name)
```

You can use multiple filters using & (and) or | (or)
Counties in the Northeast with high educational attainment:
census_data_elites<-census_data %>%

```{r}
  filter(region=="NE" & BADeg_pct > 25)
```

Counties in the Southeast or Northeast with high poverty rates:

```{r}
census_data_elites1<-census_data %>%
  filter(region=="NE" | region=="SE") %>%
  filter(BADeg_pct > 25)
table(census_data_elites1$St_name)
```

## You try it!!
Filter all counties with a poverty rate (pov_pop_pct) > 20% (BADeg_pct)
Call this census_data_pov


# Group_by, count, select, and summarise

Build a tally of "elite" counties by state 
We're using the tidyverse function "count" here to count rows by a categorical variable.


```{r}
census_data_elites_tbl<-census_data_elites %>%
  count(st_name)
```

The "select" command allows you to reduce the number of variables
You can select by column names or numbers
names(census_data)


```{r}
census_data_pov<-census_data %>%
  select(GEOID,st_name,region,pov_pop_pct)
```


Use a : to select a range of variable names.
Let's select just educational variables


```{r}
census_data_ed<-census_data %>%
  select(GEOID,st_name,region,LessHS_pct:GradDeg_pct)
```

What does that dataset look like?


```{r}
head(census_data_ed)
```

Lastly, you can use mutate to create new variables

```{r}
census_data_ed<-census_data_ed %>%
  mutate(LessBA_pct=HSGrad_pct+SomeCol_pct)
hist(census_data_ed$LessBA_pct)
boxplot(census_data_ed$LessBA_pct)
```

## You try it! 
Information about Taylor Swift and Beyonce songs on 
Spotify is available in the data folder on Github: swift_beyonce_spotify.csv
Load the data into R. Then try the following
 * Filter the dataset to look at only songs in A major (key_mode)
 * Count the number of songs by key AND artist name (hint: the `count` function can take multiple variables)
 * Create a histogram for the danceability variable for an artist of your choosing
 * Create a new variable using mutate that sums danceability, energy, and loudness
 * Calculate which artist has the highest mean value for your new variable


