---
title: "Geog4/6300: Importing data with APIs"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
While CSVs can be convenient ways to import data, APIs (application program interfaces) provide another way to access data directly from the source. Many orgs provide APIs for those using their data. 

# Working with Daymet

One example is the Daymet dataset. Created by NASA, the Daymet dataset provides gridded climate data going back to 1980 for the whole US. Here's their website: 

* https://daymet.ornl.gov/

They provide an API service where you can query their database directly. Directions are found here: 

* https://daymet.ornl.gov/web_services.html

For instance, the following uRL gives you data on the Geography/Geology building. You can create a free account with NASA to download these data.

* https://daymet.ornl.gov/data/send/query?lat=33.948693&lon=-83.375475&year=2016

The daymetr package allows you to access these data directly. We'll use this dataset in your next lab.

```{r}
#install.packages("daymetr")
library(daymetr)
```

If we wanted to download ten years of data for Athens, we can do so using the 'download_daymet' function. The value for site will be the name of the new file. 

```{r}
athens_daymet<-download_daymet(site="athens",lat=33.948693,lon=-83.375475,start=2010,end=2020,internal=TRUE)
```

This returns a list. Take a second and check it out by clicking on it in the "Environment" tab. How would you describe its structure? 

Lists don't play well with our analysis tools, so let's extract the tabular data. One of the things in this list object is something called "data", which is a data frame. That sounds promising, so let's extract that.

```{r}
athens_data<-athens_daymet$data
View(athens_data)
```

We can then begin to visualize these data.

```{r}
hist(athens_data$prcp..mm.day.)
boxplot(athens_data$tmax..deg.c.)
```

## You try it! 

* Calculate the median maximum and minimum temperatures during the time period using these Athens data. (function is `median`)
* Try downloading 10 years of climate data for a point of your choosing.

```{r}
median(athens_data$tmin..deg.c.)
median(athens_data$tmax..deg.c.)
```




# Another example of reading data from the web

The Athens-Clarke County open data website (hosted by ESRI) provides a URL for direct loading of data from their servers, so you always have the most up to date dataset. For example, we can get city park boundaries from this page: https://data-athensclarke.opendata.arcgis.com/datasets/AthensClarke::acc-parks/about. 

There's a link there to download the data as a geojson spatial data file, which we can do with the following code. Find it by clicking on "I want to use this" in the lower left and then "View API Resources". This code also uses the tmap package to map out these data.

```{r}
library(sf)


parks<-st_read("https://enigma.accgov.com/server/rest/services/ACC_Parks/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")

library(tmap)
tm_shape(parks)+
  tm_borders()

tmap_mode("view")
tm_shape(parks)+
  tm_polygons("Public_",alpha = 0.6)
```

Practice task: Find another dataset of interest on ACC's data website and load it into R.
https://data-athensclarke.opendata.arcgis.com/

```{r}

```
