---
title: "Geography 4300/6300: Reading in Data Using CSVs and Basic Manipulation"
format: html
editor_options: 
  chunk_output_type: console
---
Before you start this script, create a new project on your computer or flash drive for the course.

Projects make working with and sharing research a lot easier and are always a good practice. Move this script and all other data for the course into that folder.

You should have already installed the **tidyverse** package on your computer. Now we just need to call it.

```{r}
library(tidyverse)
```

## Download County-Level Data from GitHub

Next, download some county-level data from GitHub. You can read it in using the `read_csv()` command in the tidyverse.

```{r}
census_data <- read_csv(
  "https://github.com/communitymaplab/geog4300_datascience/raw/refs/heads/main/data/ACSCtyData_2022ACS.csv"
)
```

(Quick check--which package in the tidyverse has the `read_csv` function? Use ?read_csv on the command line to find out.)

The data page of the project site lists multiple data sources in the course GitHub repository:

<https://communitymaplab.github.io/geog4300_datascience/course_data.html>

Pick another CSV dataset and load it into R.

You have two options:

1. Right-click and save the "Raw" file link from GitHub.
2. Download the dataset to your course project folder and load it from there.

## Explore the Data

Now open the data to look at it. You can click on it in the Environment tab, or use:

```{r}
View(census_data)
```

Let's take a look at this file.

The `names()` function lists the field names.

```{r}
names(census_data)
```

The `str()` command gives you field names and data types.

```{r}
str(census_data)
```

The `head()` and `tail()` functions give you the first and last rows.

```{r}
head(census_data)

tail(census_data)
```

You can also refer to specific variables. The `$` symbol allows you to refer to them within the census frame. Let's pull median income.

Notice how this new object shows up in the Environment tab. What type of object is it?

```{r}
medinc<-census_data$medinc
```

We don't have to create a new object for this variable. We can just refer to it directly.

Let's get the mean and median values for median income, removing any NA values.

```{r}
mean(census_data$medinc,na.rm=T)

median(census_data$medinc,na.rm=T)
```

You can also create a histogram of it.

```{r}
hist(census_data$medinc)
```


## You try it!
- What are the mean and median values for the percentage of households with broadband (`intnet_brd_pct`) in each county?
- Create a histogram of that variable.

## Thinking About AI Usage

Copy the output of `str(census_data)` above and paste it into the GPT AI of your choice. Then ask it the following:

1. Create code that gives me a bar chart of the `LessHS_pct` to `GradDeg_pct` variables.
2. Explain to me how to create a bar chart of the `LessHS_pct` to `GradDeg_pct` variables, but don't give me code.
3. These are census rates computed from the American Community Survey. What are two other ways I could visualize them? Don't show me the code.
4. [Paste in the code you used above for your histogram]. How can I add a title to this plot in R?

What do you see as the potential value of all three responses?