---
title: "R Fundamentals with Daymet Climate Data"
format: html
---
[Created with assistance from Microsoft Co-pilot]

This script will give you experience working four common R objects:

- Vectors
- Lists
- Matrices
- Data frames

You will also work with several different data types:

- Character
- Numeric
- Logical
- Factor


To do this, we'll continue work with the `daymetr` package

Let's load that and the tidyverse.

```{r}
library(tidyverse)
library(daymetr)
```

# Step 1: Create Basic Vectors

Let's say we want to look at climate conditions in four Georgia cities. We can start by creating basic vectors for their names and coordinates.

```{r}
site_name <- c(
  "Athens",
  "Atlanta",
  "Savannah",
  "Augusta"
)

latitude <- c(
  33.95,
  33.75,
  32.08,
  33.47
)

longitude <- c(
  -83.38,
  -84.39,
  -81.09,
  -81.97
)
```

Take a look at those objects in the Environment tab, or use the `str` or `glimpse` functions. What types of data are each stored as?

## Practice

Create a vector named `elevation` containing four elevation values in the correct order. You can look those elevations up online.

```{r}
# your code here
```

# Step 2: Create Logical Vectors

Logical variables contain only TRUE or FALSE.

Suppose we decide that sites south of 33° latitude are considered coastal. We can run a logical test on the `latitude` object and then save its result. 

```{r}
coastal_site <- latitude < 33
coastal_site
```

Check its data type. What does "FALSE" and "TRUE" mean in this context?

## You try it!

Create a logical vector identifying sites west of -83.37 longitude. This is the longitude of Athens, so it's west or east of our city.

```{r}
# your code here
```

# Step 3: Create Factors
Factors are a form of categorical data in cases where a variable may contain repeated values. Using factors saves memory and speeds analysis, since the variable is stored using a factor ID--usually numeric like 1, 2, 3, etc.

For example, let's classify sites into regions.

```{r}
region <- factor(
  c(
    "North",
    "North",
    "Coastal",
    "Central"
  )
)
```

We can then look at that factor. How is it different from a character vector like `site_name`?

```{r}
region

levels(region)
str(region)
```

Check its structure.

```{r}
str(region)
```

## You try it!

Create a factor called `urban_class` with categories:

- Urban
- Urban
- Rural
- Urban

```{r}
# your code here
```

# Step 4: Download Daymet Data

We used `download_daymet` in a previous script. Let's download data just for Athens just to take a closer look at this list object.

```{r}
athens_weather <- download_daymet(
  site = "Athens",
  lat = 33.95,
  lon = -83.38,
  start = 2023,
  end = 2025,
  internal = TRUE
)
```

Daymet returns a list, which can contain many object types.

```{r}
str(athens_weather)
```

As we did previously, we can extract the weather data frame using the `$` symbol to refer to its name inside the list object.

```{r}
athweather_tbl <- athens_weather$data

head(athweather_tbl)
```

# Step 6: Create a function to download data for each city

We could download each city separately for analysis, but we have all the ingredients to do that at once. To start with, let's combine the separate vectors we have created into a single data frame. We can do so using the `data.frame` function, which as the name suggests creates data frames from vectors.

```{r}
ga_city<-data.frame(site_name,region,latitude,longitude)
```

Now we can create a custom function that uses this data to download climate data from Daymet:

```{r}
daymet_get<-function(site_sel){
  site_data<-ga_city %>% filter(site_name==site_sel) # Filters the dataset to just the site we specified
  
  daymet_data<-download_daymet(
      site = site_sel,
      lat = site_data$latitude,
      lon = site_data$longitude,
      start = 2025,
      end = 2025,
      internal = TRUE
  )
  
  data_tbl<-daymet_data$data %>%
    mutate(site_name=site_sel) #Add the city name to the downloaded data
}
```

Using the `map_df` function from the `purrr` package, we can download all cities at once using this function.

```{r}
city_data<-map_df(ga_city$site_name,daymet_get)
```

# Step 6: Data Frames

What data types are present in our downloaded dataset?

```{r}
str(city_data)
```

# Step 7: Creating New Variables

Create a logical variable identifying hot days.

```{r}
city_data <- city_data %>%
  mutate(
    hot_day = tmax..deg.c. > 30
  )
```

Count hot days.

```{r}
city_data %>%
  count(hot_day)
```

## You try it!

Create code that counts cold days and or wet days? Units are degrees Celsius or millimeters respectively.

# Step 8: Convert Character to Factor

Convert month values into categories.

```{r}
city_data <- city_data %>%
  mutate(
    season = case_when(
      yday < 80 ~ "Winter",
      yday < 172 ~ "Spring",
      yday < 264 ~ "Summer",
      yday < 355 ~ "Fall",
      TRUE ~ "Winter"
    ),
    season = factor(season)
  )
```

Inspect the factor.

```{r}
levels(city_data$season)
```

# Step 9: Create a Matrix

The following code uses some data wrangling to change our data into a matrix, with the rows being the day of the year (yday) and the columns being each city.

```{r}
temp_matrix <- city_data %>%
  select(site_name, yday, `tmax..deg.c.`) %>%
  pivot_wider(
    names_from = site_name,
    values_from = `tmax..deg.c.`
  ) %>%
  column_to_rownames("yday") %>%
  as.matrix()
```

Check its dimensions.

```{r}
dim(temp_matrix)
```

Find the mean value.

```{r}
mean(temp_matrix)
```

What's the mean temperature by city?

```{r}
colMeans(temp_matrix)
```

What's the mean temerature by day?

```{r}
rowMeans(temp_matrix)
```

## How is this matrix different from a data frame?