---
title: "Spatial functions and geocoding"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#install.packages(c("tidygeocoder"))

library(tidyverse)
library(tmap)
library(tidygeocoder)
library(sf)
```

# Overview

The sf package contains many spatial functions that you may be familiar with from the toolbox in ArcGIS or QGIS. These include buffers, spatial joins, dissolves, or intersections. We will also do a walkthrough of geocoding in R.

# Geocoding

Geocoding is a very common process in spatial analysis. At UGA, we have an institutional ESRI license that covers the cost of modest geocoding projects. However, there are multiple other tools available for geocoding within R. The ggmap package allows you to geocode through the Google Maps API. There is a limited cap for free geocoding and then a cost above that amount. 

For this script, we will use the free U.S. Census geocoder to map hospitals in and around Georgia. Let's read that data and get a tally of states.

```{r}
#| message: false
hosp<-read_csv("data/cms_ahd_hospitals.csv") 

table(hosp$State)
```

The tidygeocoder package provides an interface for several geocoding services. In this case, we'll use one provided by the U.S. census for just the hospitals in Georgia

```{r}
hosp_geo<-hosp %>%
  filter(State=="GA") %>%
  geocode(street=Address,city=City_Town,state=State,postalcode=ZIP_Code,method="census") 
```

Inspection of the data shows that there's several hospitals that didn't geocode at all. For now, we will filter those out and map the remaining ones. In real life, we'd have to figure out why that happened and figure out their actual coordinates.

```{r}
hosp_geo1<-hosp_geo %>%
  filter(is.na(long)==FALSE) %>%
  st_as_sf(coords=c("long","lat"),crs=4326,remove=FALSE)

tmap_mode("view")
tm_shape(hosp_geo1)+
  tm_dots()
```

***You try it!***
The ArcGIS geocoder is also available. Try geocoding the same Georgia hospitals using that one. You just need to change the method to "arcgis". How many complete entries does it give?

```{r}

```

You can read more about geocoding with tidygeocoder here: https://cran.r-project.org/web/packages/tidygeocoder/vignettes/tidygeocoder.html

# Spatial functions in R

You can also use R to perform spatial functions, like dissolves, unions, intersects, or buffers. Here's one cheat sheet with several of those functions: 

https://github.com/rstudio/cheatsheets/blob/main/sf.pdf

For example, we can buffer the hospital points. Here, we transform the data to the Georgia East State Plane projection (EPSG:6445), which uses feet as units, and draw a 5 mile buffer.

```{r}
hosp_buffer<-hosp_geo1 %>%
  st_transform(6445) %>% 
  st_buffer(dist=5280*5) #Feet/mile * number of miles

tm_shape(hosp_buffer) +
  tm_polygons()
```

# Mapping proximity
What if we want to identify the number of people in these buffers for Fulton County?

We can start by downloading a block shapefile using the tigris package, which is also made by Kyle Walker. Blocks are from the Census Tiger files.

```{r}
blocks<-tigris::blocks(state=13,county=121)

tm_shape(blocks)+
  tm_polygons()
```

Blocks are very detailed! What if we wanted to dissolve the block boundaries to create tract polygons? We can just use `group_by` and `summarise`, and sf will dissolve the boundaries.

```{r}
blocks_tract<-blocks %>%
  group_by(STATEFP20,COUNTYFP20,TRACTCE20) %>%
  summarise()

tm_shape(blocks_tract)+
  tm_polygons(border.col = "blue")+
tm_shape(st_centroid(blocks))+
  tm_dots()
```

We could also do a "difference" of those tract polygons to map those areas outside the buffers were created around hospitals. We will do so using the `ms_erase` function from `rmapshaper` (which also has an excellent online platform: https://mapshaper.org/)

```{r}
#install.packages(rmapshaper)
library(rmapshaper)

#We need the CRS projections for these two layer to match!
st_crs(hosp_buffer)

blocks_no_hosp<- blocks_tract %>%
  summarise() %>%
  st_transform(st_crs(hosp_buffer)) %>%
  ms_erase(hosp_buffer)

tm_shape(blocks_tract)+
  tm_polygons(border.alpha=0,col="#bfb90d")+
tm_shape(blocks_no_hosp)+
  tm_polygons(col="#1204b3",border.alpha=0.5)
```

**You try it!**
How many blocks are NOT within a 5 mile buffer of a hospital? Use st_centroid to create block centroids and st_join to do a spatial join to the hospital buffers. Then count the number of blocks without a joined hospital buffer.

```{r}

```

# Measuring proximity
There are several functions in R packages that allow for distance mapping. Outside of `sf`, the `nngeo` package has a function called "st_nn" that is helpful for identifying the closest neighbors by Euclidean (straight-line) distance. We will also convert the block polygons to centroids to make mapping easier (point to point is faster than polygon to point).

```{r}
#install.packages("nngeo")
library(nngeo)

blocks_hosp<-blocks %>%
  st_transform(st_crs(hosp_geo1)) %>%  #Transform blocks to use the same crs as the WH data 
  st_centroid() #change from polygons to centroids

dist_calc<-st_nn(blocks_hosp,hosp_geo1,returnDist = TRUE) #returnDist includes the distance in meters
```

Notice that the wh_dist object is a *list*, not a data frame. That list actually includes TWO dataframes. One gives the row number of the closest hospital in the `hosp_geo1` dataset. The second gives the distance in meters. Let's add the values for that second one as a new column in our block centroid dataset

```{r}
blocks_hosp$dist<-as.numeric(dist_calc$dist)

tm_shape(blocks_hosp) +
  tm_dots("dist")
```

## You try it!
What if we want to map these points as polygons rather than dots? How could we do that?
AND/OR
What if we wanted the distances to be in kilometers rather than meters?
AND/OR
Does st_nn allow you to find the distance to the THIRD closest hospital? How would you do it if so?


```{r}

```

For network/travel distances, Google and Mapbox are two of several providers whose API provides routing and travel time/distances. Both are available via R packages.

* Google example: https://hwangnyc.medium.com/calculating-travel-time-and-distance-using-google-maps-api-in-r-bbc5b74df066
* Mapbox example: https://walker-data.com/mapboxapi/articles/navigation.html
